---
title: "ASESII 24A02 Project-Based Quiz 1"
subtitle: "Ridge, Lasso, and Regularization for Neural Features"
author: |
  | Group 04
  | Nguyen Thi Yen Nhi (24125071), Nguyen Khang Thinh (24125104)
  | Nguyen Van Tinh (24125106), Nguyen Le Quoc Huy (24125100)
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
    fig_crop: false
  html_document:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
params:
  group_number: 04
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere,commandchars=\\\{\}}
  - \DefineVerbatimEnvironment{verbatim}{Verbatim}{breaklines,breakanywhere}
---

<!--
Copy this template into your group submission folder, rename it
GroupXX_ProjectQuiz1.Rmd, place references.bib beside it, and set the group
number parameter before rendering. Remove instructional comments from the
final report.
-->


# Authorship and contribution statement {-}

We confirm that every group member can explain the submitted analysis. AI use is reported in `Group04_AI_Log.csv`, and the group checked every AI-assisted item used in this report.

| Student | ID | Main contributions | Verification performed |
|---|---|---|---|
| Nguyen Thi Yen Nhi | 24125071 | [Work] | [Check] |
| Nguyen Khang Thinh | 24125104 | [Work] | [Check] |
| Nguyen Van Tinh | 24125106 | [Work] | [Check] |
| Nguyen Le Quoc Huy | 24125100 | [Work] | [Check] |

# Abstract {-}

*Maximum 150 words: prediction problem, methods, main evidence, and recommendation.*

# Reproducible setup {-}

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c("tidyverse", "glmnet", "broom", "knitr", "car")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  stop("Install the required packages before rendering: ",
       paste(missing_packages, collapse = ", "))
}

library(tidyverse)
library(glmnet)
library(broom)
library(knitr)
library(car)

# `params` exists only while knitting; the fallback
# keeps interactive chunk runs working.
has_params <- exists("params") && !is.null(params$group_number)
group_number <- if (has_params) as.integer(params$group_number) else 4L
if (length(group_number) != 1L || is.na(group_number) || group_number < 1L) {
  stop("Set params$group_number to your assigned positive group number.")
}

seeds <- list(
  split = 240200L + group_number,
  cv = 240300L + group_number,
  features = 240400L + group_number
)
```

## Shared helper functions {-}

```{r helpers}
find_exam_data <- function(filename) {
  input <- knitr::current_input(dir = TRUE)
  input_dir <- if (length(input) && nzchar(input)) dirname(input) else getwd()
  candidates <- c(
    file.path(input_dir, "data", filename),
    file.path(input_dir, "..", "data", filename),
    file.path(getwd(), "data", filename)
  )
  existing <- candidates[file.exists(candidates)]
  if (!length(existing)) stop("Cannot locate ", filename, " by a relative path.")
  hits <- unique(normalizePath(existing, winslash = "/", mustWork = TRUE))
  if (length(hits) > 1L && length(unique(unname(tools::md5sum(hits)))) > 1L) {
    stop("Multiple nonidentical copies of ", filename, " were found.")
  }
  hits[[1L]]
}

split_rows <- function(n, train_prop = 0.80, seed) {
  stopifnot(n >= 10L, train_prop > 0, train_prop < 1)
  set.seed(seed)
  train <- sort(sample.int(n, floor(train_prop * n), replace = FALSE))
  list(train = train, test = setdiff(seq_len(n), train))
}

fit_scaler <- function(x, tol = 1e-12) {
  x <- as.matrix(x)
  center <- colMeans(x)
  centered <- sweep(x, 2L, center, "-")
  scale <- sqrt(colMeans(centered^2))
  keep <- is.finite(scale) & scale > tol
  if (!any(keep)) stop("No nonconstant training predictor remains.")
  list(
    center = center[keep],
    scale = scale[keep],
    keep = keep,
    dropped = colnames(x)[!keep]
  )
}

apply_scaler <- function(x, prep) {
  x <- as.matrix(x)[, prep$keep, drop = FALSE]
  x <- sweep(x, 2L, prep$center, "-")
  sweep(x, 2L, prep$scale, "/")
}

make_foldid <- function(n, k = 5L, seed) {
  if (k < 3L || k > n) stop("Require 3 <= k <= number of training rows.")
  set.seed(seed)
  sample(rep(seq_len(k), length.out = n))
}

fit_cv_glmnet <- function(x, y, alpha, foldid) {
  glmnet::cv.glmnet(
    x = x,
    y = y,
    family = "gaussian",
    alpha = alpha,
    foldid = foldid,
    standardize = FALSE,
    intercept = TRUE,
    type.measure = "mse"
  )
}

score_regression <- function(y, prediction) {
  y <- as.numeric(y)
  prediction <- as.numeric(prediction)
  if (length(y) != length(prediction) || any(!is.finite(c(y, prediction)))) {
    stop("Metrics require equal-length finite vectors.")
  }
  c(RMSE = sqrt(mean((y - prediction)^2)),
    MAE = mean(abs(y - prediction)))
}

count_nonzero <- function(fit, s = "lambda.min", tol = 1e-8) {
  b <- as.matrix(stats::coef(fit, s = s))
  slopes <- b[rownames(b) != "(Intercept)", 1L]
  sum(abs(slopes) > tol)
}

safe_condition_numbers <- function(x, lambda, tol = 1e-10) {
  eigenvalues <- eigen(
    crossprod(x) / nrow(x),
    symmetric = TRUE,
    only.values = TRUE
  )$values
  largest <- max(eigenvalues)
  smallest <- max(min(eigenvalues), 0)
  cutoff <- tol * max(1, largest)
  c(
    gram = if (smallest <= cutoff) Inf else largest / smallest,
    ridge = (largest + lambda) / (smallest + lambda)
  )
}
```

# Problem 1. Prediction design and leakage control

## 1A. Target and predictors

```{r data-configuration}
config <- list(
  file = "prostate.csv",
  response = "lpsa",
  excluded = "lpsa"
)
data_raw <- read.csv(find_exam_data(config$file), check.names = FALSE)
predictors <- setdiff(names(data_raw), config$excluded)
analysis_data <- data_raw[, c(config$response, predictors), drop = FALSE]

if (anyNA(analysis_data)) stop("Declare a training-only missing-value plan.")
if (!all(vapply(analysis_data, is.numeric, logical(1)))) {
  stop("The assigned response and predictors must be numeric.")
}


```

**Assignment.** Group 04 is even-numbered, so the assigned dataset is
`prostate.csv` and the assigned response is `lpsa`.

**Dataset overview.** The `prostate.csv` file originates from the study by
@stamey1989, which examined the relationship between prostate-specific antigen
(PSA) levels and several clinical and histological measurements in men with
clinically localized prostate cancer who underwent radical prostatectomy. The
dataset contains `r nrow(data_raw)` observations and `r ncol(data_raw)` variables
--- one continuous response (`lpsa`) and eight candidate predictors covering
tumour characteristics, patient demographics, and biopsy summaries. All variables
are numeric and the dataset has no missing values, so no imputation or factor
encoding is required. Individual variable definitions are detailed in the overview
table below.

**Response.** `lpsa` is the natural logarithm of prostate-specific antigen
(ng/mL), a continuous serum marker measured before surgery. It is modelled on
the log scale, so every coefficient is read as a change in log PSA per one
standard deviation of a standardized predictor.

**Unit of observation and study population.** One row is one male patient
with clinically localized prostate cancer who underwent radical prostatectomy
in the study of @stamey1989. The analysis file contains `r nrow(data_raw)`
complete patients. Any conclusion therefore generalizes only to comparable
prostatectomy candidates, not to unscreened men in the general population.

**Prediction objective.** Predict pre-operative `lpsa` from routinely
available clinical and biopsy measurements, and compare how OLS, ridge, and
lasso trade prediction error against coefficient stability and interpretability
when several predictors are correlated.

**Candidate predictors.** All `r length(predictors)` remaining variables are
retained as candidate predictors. Every one is numeric, so the template's type
guard passes and the design matrix needs no factor encoding. The variables are
drawn from the clinical and pathological records described in @stamey1989
(Table 1, pp. 1077--1078) and are summarized below.

- `lcavol` --- log of cancer volume (log cm$^3$), estimated by ultrasound
  before surgery. Larger tumours are expected to produce more PSA.
- `lweight` --- log of total prostate weight (log g), measured at
  prostatectomy. Reflects overall prostate size, including both cancerous and
  non-cancerous tissue.
- `age` --- patient age at the time of diagnosis (years).
- `lbph` --- log of the amount of benign prostatic hyperplasia (log cm$^2$).
  BPH is non-cancerous prostate enlargement and can independently elevate PSA.
- `svi` --- seminal vesicle invasion (binary: 1 = present, 0 = absent),
  indicating whether cancer has spread beyond the prostate into the seminal
  vesicles. This is a known marker of advanced disease.
- `lcp` --- log of capsular penetration (log cm), measuring the extent to
  which cancer has invaded through the prostate capsule. Together with `svi`,
  it captures local disease progression.
- `gleason` --- combined Gleason score (integer, ranging from 6 to 9 in this
  sample), the standard histological grading system for prostate cancer
  aggressiveness assigned by a pathologist from biopsy tissue.
- `pgg45` --- percentage of Gleason grades 4 or 5 in the biopsy specimen
  (percent, 0--100). Higher values indicate a larger proportion of
  poorly-differentiated, aggressive tumour tissue.

All eight predictors were recorded before or during the prostatectomy procedure
and are defined independently of the serum PSA measurement, so none introduces
outcome leakage. The overview table below records the unit, data type, number
of distinct values, and missing-value count for each variable.

**Exclusions.** The project brief requires only that `lpsa` itself be removed
from the predictor set, and permits a further exclusion only when a statistical
reason is documented. We document none, so exactly one variable is excluded: the
response. No candidate predictor is computed from `lpsa`: all `r length(predictors)`
are measurements taken before surgery (tumour volume, prostate weight, age, benign
hyperplasia, seminal vesicle invasion, capsular penetration, and the two Gleason
summaries), and none uses PSA information in its definition.

**Leakage rationale.** A variable computed from the response carries the answer
inside the design matrix. Its least-squares coefficient would absorb nearly all
explained variance, the training and cross-validation errors of every method
would collapse toward zero, and the resulting model would fail on any new patient
whose PSA is not yet known --- which is precisely the deployment situation. The
comparison itself would also become meaningless: OLS, ridge, and lasso would all
be fitting the same leaked signal, so differences between them would reflect how
each penalty shrinks that one dominant column rather than how each method handles
genuine multicollinearity among clinical predictors. Because none of the eight
candidate predictors in `prostate.csv` is derived from PSA, no additional
exclusion beyond the response is warranted.

```{r data-overview}
meaning <- c(
  lcavol  = "Log cancer volume",
  lweight = "Log prostate weight",
  age     = "Age at diagnosis",
  lbph    = "Log BPH amount",
  svi     = "Seminal vesicle invasion",
  lcp     = "Log capsular penetration",
  gleason = "Gleason score",
  pgg45   = "Percent Gleason grade 4/5",
  lpsa    = "Log prostate-specific antigen"
)
unit <- c(
  lcavol = "log cm3", lweight = "log g",   age   = "years",
  lbph   = "log cm2", svi     = "0/1",     lcp   = "log cm",
  gleason = "6 to 9", pgg45   = "percent", lpsa  = "log ng/mL"
)

# A column absent from the lookups would otherwise print as a silent blank cell.
stopifnot(all(names(data_raw) %in% names(meaning)),
          all(names(data_raw) %in% names(unit)))

data_overview <- tibble(
  Variable = names(data_raw),
  Role     = ifelse(names(data_raw) == config$response, "Response", "Predictor"),
  Meaning  = unname(meaning[names(data_raw)]),
  Unit     = unname(unit[names(data_raw)]),
  Type     = vapply(data_raw, function(x) class(x)[1L], character(1)),
  Distinct = vapply(data_raw, dplyr::n_distinct, integer(1), na.rm = TRUE),
  Missing  = vapply(data_raw, function(x) sum(is.na(x)), integer(1))
)

kable(
  data_overview,
  caption = paste("Response and candidate predictors, with the integrity checks",
                  "run before the split.")
)

n_duplicate_rows <- sum(duplicated(analysis_data))
```

**Reading the table.** The file holds `r nrow(data_raw)` patients and
`r ncol(data_raw)` columns. Three integrity results clear the way for the split.
Every column is numeric, so the type guard above passes. No cell is missing
(`Missing` is zero throughout), so no imputation is performed; the missing-value
rule is still declared in 1B as a contingency, because a rule invented after
seeing the data would be a decision made with holdout information. And
`r n_duplicate_rows` rows are duplicated, so no patient is silently counted
twice, which would otherwise inflate the apparent sample size and let the same
record fall on both sides of the split. (`BPH` abbreviates benign prostatic
hyperplasia, the non-cancerous prostate enlargement recorded by `lbph`.)

The `Distinct` column carries the one substantive surprise. Two predictors are
numeric but not continuous: `svi` takes only 2 values (a binary indicator of
seminal vesicle invasion) and `gleason` only 4 (an ordinal biopsy score from 6 to
9). We keep both on their numeric scale, which is the standard `glmnet`
convention, but their coefficients must later be read with that discreteness in
mind rather than as smooth slopes: a "one standard deviation" change in a binary
variable is a convenient fiction, not a clinical intervention.

The `Unit` column exposes the second issue. The predictors live on wildly
different scales, from `pgg45` measured in percent (0 to 100) to several
variables already on a log scale. Penalized regression shrinks coefficients
toward zero, so a predictor measured in large units would be punished merely for
its units, not for being uninformative. This is precisely why the design matrix
is standardized in 1B, using centering and scaling constants computed on the
training rows only.

## 1B. Split and train-only preprocessing

```{r split-and-scale}
split <- split_rows(nrow(analysis_data), seed = seeds$split)
train_id <- split$train
test_id <- split$test
train_df <- analysis_data[train_id, ]

x_raw <- as.matrix(analysis_data[, predictors, drop = FALSE])
y_raw <- analysis_data[[config$response]]

x_prep <- fit_scaler(x_raw[train_id, , drop = FALSE])
x_train <- apply_scaler(x_raw[train_id, , drop = FALSE], x_prep)
x_test <- apply_scaler(x_raw[test_id, , drop = FALSE], x_prep)
y_train <- y_raw[train_id]
y_test <- y_raw[test_id]


foldid <- make_foldid(nrow(x_train), k = 5L, seed = seeds$cv)
sum(is.na(analysis_data))

save(x_train, y_train, x_test, y_test, analysis_data, train_id, test_id, foldid, file = "data/tmpData.RData")
```

*Report seeds, dimensions, scaling, folds, and leakage safeguards.*

## 1C. Training-data exploration

## Pairwise correlation among predictors
```{r eda-cor}
# Pairwise correlation among predictors
cor_matrix <- cor(
  train_df[, predictors],
  use = "complete.obs"
)

knitr::kable(
  round(cor_matrix, 2),
  caption = "Pairwise correlation matrix among predictors"
)
```

```{r eda-heatmap}
# Correlation heatmap
cor_matrix <- cor(train_df[, predictors], use = "complete.obs")

image(1:ncol(cor_matrix), 1:ncol(cor_matrix), cor_matrix, 
      axes = FALSE, xlab = "", ylab = "", 
      col = heat.colors(1000), main = "Correlation Heatmap of Predictors")
axis(1, at = 1:ncol(cor_matrix), labels = colnames(cor_matrix), las = 2)
axis(2, at = 1:ncol(cor_matrix), labels = colnames(cor_matrix), las = 1)
```
The pairwise correlation matrix was examined to identify potential relationships among predictors. Most predictors showed weak to moderate correlations, suggesting limited redundancy between features. However, some stronger positive correlations were observed, particularly between gleason and pgg45 (r = 0.78), lcavol and lcp (r = 0.68), and svi and lcp (r = 0.69). These relationships indicate that some predictors may share related clinical information. Therefore, Variance Inflation Factor (VIF) analysis was further performed to assess the overall multicollinearity effect in the regression model.


## Response Distribution 

```{r eda-response-dist}

hist(y_train)
summary(y_train)
```
The distribution of the response variable (lpsa) was examined using a histogram and summary statistics. The response values range from -0.43 to 5.58, with a median of 2.57 and a mean of 2.48. The mean and median are relatively close, indicating that the distribution is approximately symmetric with no strong skewness. Although some observations appear at the lower and higher ends of the range, no severe outliers are evident from the overall distribution. Therefore, no transformation was applied to the response variable before model training.

### Predictor Histogram  & Boxplot

```{r eda-hist}
## Histogram
par(mfrow = c(2,4))

for (var in predictors) {
  hist(
    train_df[[var]],
    main = paste("Distribution of", var),
    xlab = var,
    breaks = 15
  )
}

par(mfrow = c(1,1))

## Boxplot
par(mfrow = c(2,4))
for (var in predictors) {
  boxplot(
    train_df[[var]],
    main = var
  )
}
par(mfrow = c(1,1))


```
The histograms show that most predictors have different distribution patterns. lcavol, lweight, and age are approximately symmetric, while lbph, lcp, and pgg45 exhibit right-skewed distributions with longer tails toward higher values. svi and gleason are discrete variables with highly concentrated values. These distribution characteristics are considered in later regression analysis and model diagnostics.

Upon inspecting the visualizations, a potential influential observation is evident in the lweight vs lpsa plot. There is a single data point located at the far right with an exceptionally high lweight (exceeding 6.0) but a relatively low lpsa (approximately 2.0).

### Correlation between predictors and response

```{r eda-cor-res-pred}
cor_response <- cor(
  train_df[, predictors],
  train_df$lpsa
)

knitr::kable(
  round(cor_response, 2),
  caption = "Correlation between predictors and lpsa"
)
```

The correlation analysis indicates that all predictors have positive correlations with lpsa. lcavol shows the strongest relationship (r = 0.77), followed by lcp (r = 0.59) and svi (r = 0.57). Variables such as pgg45, gleason, and lweight have moderate correlations, while age and lbph show weak associations. Therefore, lcavol, lcp, and svi are likely to be important predictors in the multiple linear regression model.

###  Check multicollinearity using VIF
```{r eda-vif}

vif_model <- lm(
  lpsa ~ .,
  data = train_df[, c("lpsa", predictors)]
)

vif_values <- vif(vif_model)

knitr::kable(
  round(vif_values, 2),
  caption = "Variance Inflation Factor (VIF)"
)
```

There is no significant multicollinearity present in this model. Because all VIF values are safely below 5, the independent variables are not highly correlated with one another. The regression model's coefficient estimates are mathematically stable, and no variables need to be dropped or transformed for multicollinearity reasons.

###  Predictor-response plots
```{r eda-pre-res-plot}

par(mfrow = c(2,4))

for (var in predictors) {
  plot(
    train_df[[var]],
    train_df$lpsa,
    xlab = var,
    ylab = "lpsa",
    main = paste(var, "vs lpsa"),
    pch = 19
  )

  abline(
    lm(train_df$lpsa ~ train_df[[var]]),
    col = "red"
  )
}

par(mfrow = c(1,1))
```
The provided scatter plots visualize the marginal relationships between the eight candidate predictors and the target variable, lpsa.

Strong Trends: lcavol demonstrates the clearest and most distinct positive linear correlation with lpsa, suggesting it will be a strong predictor in the model.

Categorical Patterns: Variables such as svi (binary) and gleason (ordinal) display expected clustered structures, with higher categories generally corresponding to higher lpsa values.

Weak Trends: age and lbph exhibit highly dispersed data clouds with very flat regression lines, indicating weak independent linear relationships with the target.



#### Question
Given the varying individual predictive strengths of these features and the moderate structural correlations (e.g., between pgg45 and lcp), will an $L_{1}$ penalty (Lasso) completely eliminate the structurally weaker predictors like age to produce a strictly sparse model, or will an $L_{2}$ penalty (Ridge) prove more stable against the identified high-leverage anomaly in lweight?

# Problem 2. OLS, ridge, and lasso

## 2A. OLS baseline

```{r ols}
# Convert x_train (matrix) and y_train into a data frame for lm()
train_data <- data.frame(y = y_train, x_train)

# 1. Fit OLS on the training set
ols_fit <- lm(y ~ ., data = train_data)

# 2. Extract coefficients (tidy table via broom)
ols_coefs <- broom::tidy(ols_fit)
kable(
  ols_coefs,
  digits = 4,
  caption = "OLS coefficient estimates on standardized training data."
)

# 3. R-squared and Adjusted R-squared from the model summary
ols_summary <- summary(ols_fit)
cat(sprintf("R-squared:          %.4f\n", ols_summary$r.squared))
cat(sprintf("Adjusted R-squared: %.4f\n", ols_summary$adj.r.squared))

# 4. Predict on the training set and evaluate training error
ols_train_pred <- predict(ols_fit, newdata = train_data)
ols_train_metrics <- score_regression(y = y_train, prediction = ols_train_pred)
cat(sprintf("Training RMSE:      %.4f\n", ols_train_metrics["RMSE"]))
cat(sprintf("Training MAE:       %.4f\n", ols_train_metrics["MAE"]))

# 5. Five-fold CV error using the shared foldid with ridge/lasso
#    (lm() has no built-in CV, so we compute it manually)
K <- max(foldid)
cv_mse_folds <- numeric(K)
for (k in seq_len(K)) {
  idx_train_k <- which(foldid != k)
  idx_val_k   <- which(foldid == k)
  fold_data   <- data.frame(y = y_train[idx_train_k], x_train[idx_train_k, , drop = FALSE])
  fold_fit    <- lm(y ~ ., data = fold_data)
  fold_pred   <- predict(fold_fit,
                         newdata = data.frame(x_train[idx_val_k, , drop = FALSE]))
  cv_mse_folds[k] <- mean((y_train[idx_val_k] - fold_pred)^2)
}
ols_cv_mse  <- mean(cv_mse_folds)
ols_cv_rmse <- sqrt(ols_cv_mse)
cat(sprintf("5-fold CV MSE:      %.4f\n", ols_cv_mse))
cat(sprintf("5-fold CV RMSE:     %.4f\n", ols_cv_rmse))
```

```{r ols-diagnostic, fig.cap="Cook's distance for the OLS fit. Observations exceeding the 4/n threshold (dashed line) are flagged as influential.", fig.height=4}
# 6. Residual / influence diagnostic: Cook's distance
cooks_d   <- cooks.distance(ols_fit)
n_train   <- nrow(train_data)
threshold <- 4 / n_train

plot(cooks_d,
     type = "h", lwd = 1.5, col = "steelblue",
     ylab = "Cook's distance", xlab = "Observation index",
     main = "Cook's Distance — OLS Baseline")
abline(h = threshold, lty = 2, col = "firebrick", lwd = 1.5)
text_idx <- which(cooks_d > threshold)
if (length(text_idx)) {
  text(text_idx, cooks_d[text_idx],
       labels = text_idx, pos = 3, cex = 0.8, col = "firebrick")
}
cat(sprintf("Threshold (4/n):    %.4f\n", threshold))
cat(sprintf("Influential points: %d of %d\n", length(text_idx), n_train))
```

```{r ols-helpers, echo=FALSE}
# Helper functions to extract coefficient and p-value by term name
ols_beta <- function(term) ols_coefs$estimate[ols_coefs$term == term]
ols_pval <- function(term) ols_coefs$p.value[ols_coefs$term == term]

# Compute lcavol–lcp correlation on training data (used in narrative)
cor_lcavol_lcp <- cor(x_train[, "lcavol"], x_train[, "lcp"])

# Identify non-significant predictors (p > 0.05, excluding intercept)
nonsig_terms <- ols_coefs %>%
  filter(term != "(Intercept)", p.value > 0.05) %>%
  pull(term)
nonsig_max_t <- ols_coefs %>%
  filter(term %in% nonsig_terms) %>%
  summarise(max_abs_t = max(abs(statistic))) %>%
  pull(max_abs_t)
nonsig_min_p <- ols_coefs %>%
  filter(term %in% nonsig_terms) %>%
  summarise(min_p = min(p.value)) %>%
  pull(min_p)
```

**Model summary.** The OLS model is fitted on the `r nrow(x_train)`
standardized training observations with all `r ncol(x_train)` predictors.
Because every predictor has been centred and scaled to unit (population)
variance, the intercept equals the mean of `lpsa` in the training set, and
each slope measures how much `lpsa` changes per one population-SD increase in
that predictor, holding the remaining seven fixed.

**Coefficient interpretation.** Two predictors dominate.

- `lcavol` carries the largest standardized coefficient
  ($\hat{\beta} \approx `r round(ols_beta("lcavol"), 2)`$,
  $p `r ifelse(ols_pval("lcavol") < 0.001, "< 0.001", sprintf("\\approx %.3f", ols_pval("lcavol")))`$),
  confirming that log cancer volume is the single strongest linear predictor
  of log PSA.
- `svi` is the second-largest contributor
  ($\hat{\beta} \approx `r round(ols_beta("svi"), 2)`$,
  $p \approx `r round(ols_pval("svi"), 3)`$), indicating that seminal
  vesicle invasion is associated with meaningfully higher PSA even after
  adjusting for tumour volume.
- `lweight` is marginally significant
  ($p \approx `r round(ols_pval("lweight"), 3)`$); the remaining
  `r length(nonsig_terms)` predictors
  (`r paste0("\x60", nonsig_terms, "\x60", collapse = ", ")`) all have
  $|t| < `r round(nonsig_max_t + 0.05, 1)`$ and
  $p > `r round(nonsig_min_p - 0.01, 2)`$.

**Collinearity concern.** The near-zero and sign-reversed coefficient for
`lcp` ($\hat{\beta} \approx `r round(ols_beta("lcp"), 2)`$) is noteworthy:
biologically, greater capsular penetration should correlate positively with
PSA, and indeed the marginal correlation is positive. The sign flip in OLS is
a classic symptom of multicollinearity --- `lcavol` and `lcp` are strongly
correlated ($r \approx `r round(cor_lcavol_lcp, 2)`$ on the training set),
so OLS distributes their shared variance unevenly. Regularization in 2B--2C
is expected to stabilize these shared-variance coefficients.

**Cross-validation error.** The five-fold CV RMSE is computed using the same
`foldid` assignment that will be passed to `cv.glmnet` for ridge and lasso,
ensuring a like-for-like comparison. Because OLS has no tuning parameter, there
is a single CV-RMSE value rather than a curve. The gap between training RMSE and
CV RMSE reflects the optimism of the in-sample fit: a large gap would signal
overfitting, while a small gap would suggest the eight-predictor model
generalizes reasonably well given the available sample.

**Influence diagnostic.** The Cook's distance plot flags observations whose
removal would substantially shift the fitted coefficients. The conventional
threshold $4/n$ is shown as a dashed line. Any point above this line has
outsized leverage, large residual, or both. In a sample of only
`r nrow(x_train)` training rows, even one or two influential points can
destabilize an unpenalized fit --- particularly for the collinear predictors
`lcavol`/`lcp` whose coefficient estimates are already fragile. These
influential observations are retained in the analysis (not removed) because
they are clinically valid patients; their effect is instead mitigated by the
regularization introduced in 2B--2C.

## 2B. Ridge

### Cross-validation curve:

```{r ridge}

ridge <- fit_cv_glmnet(x = x_train, y = y_train, alpha=0, foldid = foldid)
plot(ridge)

```
The mean cross-validation MSE decreases as $-Log(\lambda)$ increases (equivalently, as $\lambda$ decreases), suggesting that reducing the regularization strength improves predictive performance. The left dashed line indicates $\lambda_{1se}$, which selects a more regularized model with prediction error within one standard error of the minimum, while the right dashed line indicates $\lambda_{min}$, which yields the lowest cross-validation error.

### Lambda min:
```{r}

g <- crossprod(x_train)
eigen <- eigen(g)$value

lambda <- ridge$lambda.min
cat("Lambda min: ",lambda, "\n")
cat("MSE: ", ridge$cvm[ridge$lambda == lambda], "\n")
cat("Condition number:", (max(eigen) + lambda)/(min(eigen) + lambda), "\n")

```
With $\lambda_{min} \approx 0.23$, the 5-fold cross-validation MSE reaches its minimum value of $MSE_{min} \approx 0.61$, indicating the best predictive performance among the candidate values of $\lambda$. However, the corresponding Ridge model still has a relatively large condition number, suggesting that the model remains sensitive to multicollinearity and may not be sufficiently stable.

### Lambda 1se:
```{r}

g <- crossprod(x_train)
eigen <- eigen(g)$value

lambda <- ridge$lambda.1se
cat("Lambda 1se: ",lambda, "\n")
cat("MSE: ", ridge$cvm[ridge$lambda == lambda], "\n")
cat("Condition number:", (max(eigen) + lambda)/(min(eigen) + lambda), "\n")

```
Although the one-standard-error rule selected a larger penalty $\lambda_{1se} \approx 1.02$, the condition number decreased only slightly from 20.2 to 19.07, indicating minimal improvement in numerical stability. In contrast, the cross-validation MSE increased from 0.61 to 0.66. Therefore, the additional regularization provided by $\lambda_{1se}$ offers little practical benefit while reducing predictive performance.

### Coefficients:
```{r}

data.frame(
  OLS = as.numeric(coef(ols_fit)),
  Lambda_Min = as.numeric(coef(ridge, "lambda.min")),
  Lambda_Max = as.numeric(coef(ridge, "lambda.1se")),
  row.names = names(ols_fit$coefficients)
)
```

Compared with the OLS model, Ridge regression shrinks most regression coefficients toward zero, reducing their magnitudes through the regularization penalty. This shrinkage helps alleviate the effects of multicollinearity while retaining all predictors in the model. Most coefficients exhibit the expected behavior by moving closer to zero as the regularization strength increases. However, two coefficients show notable exceptions. The coefficient of \textbf{lcp} changes from a small negative value in the OLS model to positive values under Ridge regression. In addition, the coefficient of \textbf{gleason} increases in magnitude, moving further away from zero rather than shrinking like the other coefficients. These changes suggest that the effects of \textbf{lcp} and \textbf{gleason} are influenced by multicollinearity with other predictors. As Ridge simultaneously estimates all coefficients under a penalty, it redistributes the effects among correlated variables, causing some coefficients to increase in magnitude or even change sign despite the overall shrinkage trend.

```{R}
n_vars <- nrow(ridge$glmnet.fit$beta)
plot(ridge$glmnet.fit)
legend("topleft", 
       legend = rownames(ridge$glmnet.fit$beta), 
       col = 1:n_vars, 
       lty = 1, 
       cex = 0.6)
```
As $\lambda$ increases, all regression coefficients are progressively shrunk toward zero, demonstrating the regularization effect of Ridge regression. Conversely, as $\lambda$ decreases, the coefficients gradually move away from zero toward their OLS estimates. The coefficient of \textbf{lcavol} remains the largest throughout the regularization path, indicating that it has the strongest association with the response variable. The coefficients of  \textbf{svi} and lweight also increase substantially as the regularization weakens, suggesting relatively strong effects on the response. In contrast, the coefficients of \textbf{gleason}, \textbf{lbph}, and \textbf{pgg45} remain relatively small across the entire path, indicating weaker contributions to the model. The coefficient of age remains negative throughout the regularization path, reflecting a consistent negative relationship with the response variable. Notably, the coefficient of  \textbf{lcp} exhibits non-monotonic behavior, first increasing and then decreasing as $\lambda$ changes, suggesting that its estimated effect is sensitive to multicollinearity with other predictors.

To address multicollinearity, Ridge impose an $L_2$ penalty on the regression coefficients. Rather than assigning a large coefficient to one correlated predictor and a small coefficient to another, Ridge shrinks the coefficients simultaneously and redistributes their effects among the correlated variables. Consequently, the coefficient estimates become more stable and less sensitive to small changes in the data.

## 2C. Lasso


### Model fitting and cross-validation


```{r lasso-fit}
lasso_cv <- fit_cv_glmnet(
    x = x_train,
    y = y_train,
    alpha = 1,
    foldid = foldid
)
plot(lasso_cv)

```

Based on the 5-fold cross-validation curve generated for the Lasso regression model ($\alpha = 1$), we observe the relationship between the tuning parameter $\lambda$ and the Mean-Squared Error (MSE). The top axis indicates the number of active (non-zero) predictors retained in the model at each level of penalization.

* The Cross-Validation Curve: The red dots represent the average CV-MSE across the 5 folds for each $\lambda$ value, with the gray error bars denoting one standard error above and below the mean. As $-\log(\lambda)$ increases (moving from left to right, meaning the penalty weakens), the MSE initially drops, reaches a minimum, and then slightly stabilizes as the model approaches the unpenalized OLS limit.

* Optimal $\lambda$ ($\lambda_{\text{min}}$): The right vertical dashed line marks `lambda.min`, the value that strictly minimizes the cross-validation MSE. At this optimal point for prediction accuracy ($-\log(\lambda) \approx 2.4$), the Lasso penalty retains 5 non-zero predictors*.

* Parsimonious $\lambda$ ($\lambda_{\text{1se}}$): The left vertical dashed line marks `lambda.1se`, selected by the one-standard-error rule. This value applies a stronger penalty ($-\log(\lambda) \approx 1.67$) to produce a sparser model. It retains only 3 non-zero predictors while keeping the error within one standard deviation of the absolute minimum MSE, offering a robust trade-off between predictive power and model simplicity.

### Analysis of $\lambda_{\text{min}}$

```{r lasso-min}
lambda_min <- lasso_cv$lambda.min
coef(lasso_cv, s="lambda.min")
count_nonzero(lasso_cv, s="lambda.min")
```


The optimal tuning parameter selected by five-fold cross-validation is $\lambda_{\text{min}} = 0.0891$. This value minimizes the cross-validation Mean Squared Error (CV-MSE), providing the best predictive performance among all candidate penalty values.

At this level of regularization, the Lasso model retains five non-zero predictors: lcavol, lweight, lbph, svi, and pgg45. The remaining predictors (age, lcp, and gleason) are shrunk exactly to zero, demonstrating Lasso's ability to perform automatic variable selection while maintaining predictive accuracy.

The retained predictors suggest that tumor volume (lcavol) is the most influential variable, having the largest estimated coefficient (0.6761). Variables related to body weight (lweight), benign prostatic hyperplasia (lbph), seminal vesicle invasion (svi), and the percentage of Gleason grades 4 or 5 (pgg45) also contribute to predicting PSA levels, although with relatively smaller effects.

Overall, $\lambda_{\text{min}}$ produces the model with the lowest cross-validation error while preserving a moderate number of predictors, offering a balance between prediction accuracy and model complexity.

```{r lasso-cv-error-min}
cv_mse_min <- lasso_cv$cvm[
    lasso_cv$lambda==
    lasso_cv$lambda.min
]
cv_mse_min
```
The minimum five-fold cross-validation Mean Squared Error (CV-MSE) achieved by the Lasso model is **`r round(cv_mse_min, 4)`**. This value corresponds to $\lambda_{\text{min}} = `r round(lasso_cv$lambda.min, 4)`$ and represents the estimated prediction error obtained during model tuning using the training data.

A lower CV-MSE indicates better expected predictive performance on unseen data. Since $\lambda_{\text{min}}$ is selected to minimize this metric, it is the tuning parameter that prioritizes prediction accuracy over model simplicity.


### Analysis of $\lambda_{\text{1se}}$

```{r lasso-1se}
lambda_1se <- lasso_cv$lambda.1se
coef(lasso_cv, s="lambda.1se")
count_nonzero(lasso_cv, s="lambda.1se")
```
The one-standard-error rule selects $\lambda_{\text{1se}} = r round(lambda_1se, 4)$, which applies a stronger regularization penalty than $\lambda_{\text{min}}$. The corresponding five-fold cross-validation Mean Squared Error (CV-MSE) is r round(cv_mse_1se, 4), remaining within one standard error of the minimum while producing a simpler model.

Under $\lambda_{\text{1se}}$, the Lasso model retains only three non-zero predictors (lcavol, lweight, and svi). Compared with $\lambda_{\text{min}}$, the variables lbph and pgg45 are removed, resulting in a more parsimonious and interpretable model with only a small loss in predictive performance.

```{r lasso-cv-error-1se}
cv_mse_1se <- lasso_cv$cvm[
    lasso_cv$lambda==
    lasso_cv$lambda.1se
]
cv_mse_1se
```
The one-standard-error rule selects $\lambda_{\text{1se}} = r round(lambda_1se, 4)$, resulting in a five-fold cross-validation Mean Squared Error (CV-MSE) of r round(cv_mse_1se, 4). Although this error is slightly higher than that of $\lambda_{\text{min}}$, it remains within one standard error of the minimum while providing a more conservative model.

Under $\lambda_{\text{1se}}$, the Lasso model retains only three non-zero predictors (lcavol, lweight, and svi). Compared with $\lambda_{\text{min}}$, the predictors lbph and pgg45 are removed, resulting in a simpler and more interpretable model with only a modest reduction in predictive performance.

### Coefficient Path

```{r lasso-coef-path}
lasso_fit <- glmnet(
    x_train,
    y_train,
    alpha=1,
    standardize=FALSE,
      
)

plot(lasso_fit, xvar = "lambda", label = TRUE)

legend("topleft", 
       legend = rownames(lasso_fit$beta), 
       col = 1:nrow(lasso_fit$beta), 
       lty = 1, 
       cex = 0.7,
       lwd = 2,
       title = "Predictors")

```


Overall, the coefficient path shows that as $-\log(\lambda)$ increases (equivalently, as $\lambda$ decreases), the regularization penalty becomes weaker, allowing more predictors to enter the model and their coefficients to increase in magnitude. Under strong regularization, many coefficients are shrunk exactly to zero, while weaker regularization gradually recovers the full model.

Among the predictors, **lcavol** enters the model first and consistently has the largest coefficient, indicating that it is the most influential predictor of **lpsa**. **svi** and **lweight** also become active relatively early and maintain positive coefficients throughout the path. In contrast, **age, lbph, lcp, gleason,** and **pgg45** enter only when the penalty becomes sufficiently small, suggesting that their contributions are comparatively weaker and less stable.

## 2D 

```{r core-comparison}
# Build a comparison from training/CV quantities only.
```

**Core model nominated before holdout evaluation:** [method and reason]

## 2E. Perturbation or stability check

**Prediction before running the check:** [prediction]

```{r perturbation}
# Implement one allowed training-only stability check.
```

**Observed result and explanation:** [result]

# Problem 3. Mathematical mechanisms

Let

$$
X \in \mathbb{R}^{n\times p}
$$

be the standardized training design matrix with $n$ observations and $p$ predictors.

Let

$$
y\in\mathbb{R}^{n}
$$

be the response vector and

$$
b\in\mathbb{R}^{p}
$$

be the coefficient vector.

## 3A. Ridge and conditioning

*Write the ridge derivation. Define dimensions and explain uniqueness.*

```{r ridge-conditioning}
# Use x_train and the fitted ridge lambda.min.
# safe_condition_numbers(x_train, lambda = ...)
```

## 3A. Ridge and conditioning

### Ridge objective

The ridge objective function is

$$
Q_\lambda(b)
=
\frac1{2n}\|y-Xb\|^2
+
\frac{\lambda}{2}\|b\|^2.
$$

Taking the gradient,

$$
\nabla Q_\lambda(b)
=
\frac1n(X^TXb-X^Ty)
+
\lambda b.
$$

Setting the gradient equal to zero,

$$
\frac1nX^TXb
-
\frac1nX^Ty
+
\lambda b
=
0.
$$

Rearranging,

$$
\left(
\frac1nX^TX
+
\lambda I
\right)b
=
\frac1nX^Ty.
$$

Define

$$
G=\frac1nX^TX,
\qquad
z=\frac1nX^Ty.
$$

Hence,

$$
(G+\lambda I)b=z,
$$

and therefore

$$
\boxed{
\hat b
=
(G+\lambda I)^{-1}z.
}
$$

Since \(G\) is positive semidefinite and \(\lambda>0\),

$$
G+\lambda I
$$

is positive definite and therefore invertible. Hence the ridge estimator is unique.

The spectral condition number is

$$
\kappa_2(G)
=
\frac{\mu_{\max}}
{\mu_{\min}},
$$

while ridge gives

$$
\kappa_2(G+\lambda I)
=
\frac{\mu_{\max}+\lambda}
{\mu_{\min}+\lambda}.
$$

where

$$
\mu_{\max}=\max_i\mu_i
$$

is the **largest eigenvalue** of the Gram matrix \(G\), and

$$
\mu_{\min}=\min_i\mu_i
$$

is the **smallest (positive) eigenvalue** of \(G\).

---

### Condition Number and Numerical Stability

The Gram matrix is defined as

$$
G=\frac1nX^TX.
$$

Since \(G\) is symmetric and positive semidefinite, all of its eigenvalues are non-negative. Let the eigenvalues of \(G\) be

$$
\mu_1,\mu_2,\ldots,\mu_p,
$$

where

$$
\mu_{\max}=\max_i\mu_i,
\qquad
\mu_{\min}=\min_i\mu_i.
$$

The spectral condition number is defined as

$$
\kappa_2(G)
=
\frac{\mu_{\max}}
{\mu_{\min}}.
$$

This quantity measures how sensitive the solution of a linear system is to small perturbations in the input data. A large condition number indicates that the Gram matrix is close to singular, so the estimated coefficients may change dramatically even when the training data are only slightly perturbed. Conversely, a small condition number indicates that the matrix is well-conditioned and the regression coefficients are numerically stable.

Ridge regression replaces the Gram matrix by

$$
G+\lambda I.
$$

If the eigenvalues of \(G\) are

$$
\mu_1,\mu_2,\ldots,\mu_p,
$$

then the eigenvalues of the ridge matrix become

$$
\mu_1+\lambda,\,
\mu_2+\lambda,\,
\ldots,\,
\mu_p+\lambda.
$$

Hence, the condition number becomes

$$
\kappa_2(G+\lambda I)
=
\frac{\mu_{\max}+\lambda}
{\mu_{\min}+\lambda}.
$$

Because the same positive constant is added to every eigenvalue, the smallest eigenvalue is shifted away from zero, reducing the condition number. Consequently, the matrix inversion becomes more numerically stable, the effect of multicollinearity is reduced, and the ridge estimator is guaranteed to have a unique solution.

From the table above, the ridge matrix should have a smaller condition number than the original Gram matrix. This demonstrates that ridge regularization improves the numerical conditioning of the optimization problem, leading to more stable coefficient estimates.
## 3B. Lasso optimality and soft thresholding

*Derive the zero/nonzero optimality conditions and the orthonormal-design formula.*

## 3B. Lasso optimality and soft thresholding

Lasso regression estimates the regression coefficients by minimizing

$$
Q_\lambda(b)
=
\frac1{2n}\|y-Xb\|^2
+
\lambda\|b\|_1,
$$

where

$$
\|b\|_1
=
\sum_{j=1}^{p}|b_j|,
$$

and \(\lambda>0\) controls the amount of regularization.

Unlike ridge regression, the \(L_1\)-norm contains the absolute value function, which is not differentiable at zero. Consequently, the ordinary gradient cannot be applied, and the optimal solution must be derived using the **subgradient**.

### Step 1. Rewrite the Objective Function

Assume that the predictors have been centered, standardized, and satisfy the orthonormal design condition

$$
\frac1nX^TX=I.
$$

Equivalently,

$$
X^TX=nI.
$$

Define

$$
z=\frac1nX^Ty.
$$

The squared error term can be expanded as

$$
\begin{aligned}
\|y-Xb\|^2
&= (y-Xb)^T(y-Xb)\\
&= y^Ty - 2b^TX^Ty + b^TX^TXb.
\end{aligned}
$$

Substituting into the objective function,

$$
\begin{aligned}
Q_\lambda(b)
&=
\frac1{2n}
\left(
y^Ty
-
2b^TX^Ty
+
b^TX^TXb
\right)
+
\lambda\sum_{j=1}^{p}|b_j|.
\end{aligned}
$$

Using

$$
X^TX=nI
$$

and

$$
z=\frac1nX^Ty,
$$

the objective becomes

$$
Q_\lambda(b)
=
\frac12b^Tb
-
b^Tz
+
\lambda\sum_{j=1}^{p}|b_j|
+
C,
$$

where

$$
C=\frac1{2n}y^Ty
$$

is a constant independent of \(b\).

Expanding by coordinates,

$$
Q_\lambda(b)
=
\sum_{j=1}^{p}
\left[
\frac12b_j^2
-
z_jb_j
+
\lambda|b_j|
\right]
+
C.
$$

Therefore, each coefficient can be optimized independently.


### Step 2. Compute the Subgradient

Since

$$
|b_j|
$$

is not differentiable at zero, its subgradient is

$$
\partial|b_j|
=
\begin{cases}
1,&b_j>0,\\
[-1,1],&b_j=0,\\
-1,&b_j<0.
\end{cases}
$$

The first-order optimality condition becomes

$$
0
\in
b_j-z_j
+
\lambda\partial|b_j|.
$$

The solution is obtained by considering three cases.


#### Case 1. Positive coefficient

Suppose

$$
b_j>0.
$$

Then

$$
\partial|b_j|=1,
$$

and

$$
b_j-z_j+\lambda=0.
$$

Hence

$$
b_j=z_j-\lambda.
$$

For this solution to remain positive,

$$
z_j>\lambda.
$$



#### Case 2. Negative coefficient

Suppose

$$
b_j<0.
$$

Then

$$
\partial|b_j|=-1,
$$

and

$$
b_j-z_j-\lambda=0.
$$

Therefore,

$$
b_j=z_j+\lambda.
$$

Since the coefficient must remain negative,

$$
z_j<-\lambda.
$$


#### Case 3. Zero coefficient

Suppose

$$
b_j=0.
$$

The subgradient condition becomes

$$
0
\in
-z_j+\lambda[-1,1].
$$

This is satisfied whenever

$$
|z_j|\le\lambda.
$$

Therefore,

$$
b_j=0.
$$

This explains why lasso can produce exact zero coefficients.



### Step 3. Soft-Thresholding Solution

Combining the three cases,

$$
\hat b_j=
\begin{cases}
z_j-\lambda,&z_j>\lambda,\\
0,&|z_j|\le\lambda,\\
z_j+\lambda,&z_j<-\lambda.
\end{cases}
$$

The piecewise solution can be written compactly as

$$
\boxed{
\hat b_j
=
\operatorname{sign}(z_j)
\left(|z_j|-\lambda\right)_+
}
$$

where

$$
(a)_+=\max(a,0).
$$

This expression is known as the **soft-thresholding operator**.

---

### Interpretation

The soft-thresholding operator subtracts the penalty \(\lambda\) from the magnitude of every coefficient. When

$$
|z_j|\le\lambda,
$$

the coefficient is shrunk completely to zero, meaning that the corresponding predictor is removed from the model. Therefore, lasso simultaneously performs coefficient estimation and automatic variable selection.

In contrast, ridge regression continuously shrinks coefficients toward zero,

$$
\hat b_{\text{ridge}}
=
(G+\lambda I)^{-1}z,
$$

but almost never makes them exactly zero. Consequently, ridge improves numerical stability, whereas lasso additionally produces sparse and more interpretable models.

---

### Interpretation

The soft-thresholding operator subtracts the penalty \(\lambda\) from the magnitude of every coefficient. When

$$
|z_j|\le\lambda,
$$

the coefficient is shrunk completely to zero, meaning that the corresponding predictor is removed from the model. Therefore, lasso simultaneously performs coefficient estimation and automatic variable selection.

In contrast, ridge regression continuously shrinks coefficients toward zero,

$$
\hat b_{\text{ridge}}
=
(G+\lambda I)^{-1}z,
$$

but almost never makes them exactly zero. Consequently, ridge improves numerical stability, whereas lasso additionally produces sparse and more interpretable models.


## 3C. Connect algebra to evidence

*Connect one specific numerical result to Problem 3A or 3B.*

ridge$lambda.min

G <- crossprod(x_train) / nrow(x_train)

eig <- eigen(G)$values

lambda <- ridge$lambda.min

kappa_before <- max(eig) / min(eig)

kappa_after <- (max(eig) + lambda) /
               (min(eig) + lambda)

data.frame(
  Matrix = c("Gram matrix (G)", "Ridge matrix (G + λI)"),
  Condition_Number = c(kappa_before, kappa_after)
)

## 3C. Connect algebra to evidence

Section 3A showed that ridge regression improves the conditioning of the Gram matrix by adding a positive constant \(\lambda\) to every eigenvalue. Consequently, the theoretical condition number changes from

$$
\kappa_2(G)
=
\frac{\mu_{\max}}
{\mu_{\min}}
$$

to

$$
\kappa_2(G+\lambda I)
=
\frac{\mu_{\max}+\lambda}
{\mu_{\min}+\lambda}.
$$

To verify this result, we computed the condition numbers of the original Gram matrix and the ridge-regularized matrix using the optimal tuning parameter selected by cross-validation.

```{r ridge-evidence}
G <- crossprod(x_train) / nrow(x_train)

eig <- eigen(G)$values

lambda <- ridge$lambda.min

kappa_before <- max(eig) / min(eig)

kappa_after <- (max(eig) + lambda) /
               (min(eig) + lambda)

knitr::kable(
  data.frame(
    Matrix = c("Gram matrix (G)", "Ridge matrix (G + λI)"),
    Condition_Number = c(kappa_before, kappa_after)
  ),
  digits = 3,
  caption = "Condition numbers before and after ridge regularization."
)
```

The results show that the condition number decreases from **20.596** for the original Gram matrix to **8.936** after ridge regularization. This agrees with the theoretical derivation in Section 3A, confirming that adding the penalty term shifts every eigenvalue away from zero and improves the conditioning of the optimization problem. Consequently, the matrix inversion becomes more numerically stable, the effect of multicollinearity is reduced, and the estimated regression coefficients become more reliable.

# Problem 4. Elastic net and a neural-feature bridge

## 4A. Research question and source map

**Research direction:** [weight decay / sparse weights / pruning / dropout]

| Claim | Exact primary source and locator | What it establishes | What it does not establish |
|---|---|---|---|
| [Claim] | [Citation, section/page/equation] | [Support] | [Boundary] |

## 4B. Elastic net on original predictors

```{r original-elastic-net}
alpha_grid <- c(0.10, 0.25, 0.50, 0.75, 0.90)

# Fit one cv.glmnet object per alpha with the same foldid.
# Select alpha and lambda using training CV only.
```

## 4C. Fixed ReLU features

```{r random-relu-features}
relu <- function(z) pmax(z, 0)
M <- 200L

# TODO:
# 1. Generate A and bias once from seeds$features using the stated distributions.
# 2. Apply the same A and bias to x_train and x_test.
# 3. Fit a scaler to H_train only and apply it to H_test.
# 4. Fit ridge, lasso, and elastic net on H_train with the shared foldid.
```

*State which weights are fixed, which are learned, and how active features are counted.*

## 4D. Locked declaration and final holdout evaluation

> **Locked before viewing `y_test`:** [preprocessing, candidate specifications, selected tuning values, feature seed, nominated deployment model, metrics]

```{r final-holdout-evaluation}
# This must be the first analysis chunk that uses y_test for evaluation.
# Generate all fixed candidate predictions and one RMSE/MAE table here.
```

*State whether the holdout evidence supports the predeclared choice. Do not retune.*

## 4E. Deep-learning connection and limitations

*Distinguish penalty, weight decay, output-weight sparsity, pruning, dropout, and a trained deep network. Cite the sources from 4A and discuss at least two limitations.*

# Problem 5. Report quality and reproducibility

## 5A. Reproduction record

```text
1. Open a clean R session.
2. [Exact commands and folder layout.]
3. Render GroupXX_ProjectQuiz1.Rmd.
```

## 5B. Recommendation and limitations

*Give the final recommendation, distinguish prediction from interpretation, and state limitations.*

## 5C. AI verification record

| AI-assisted item | How it was checked | Evidence | Modification or rejection |
|---|---|---|---|
| [Item] | [Method] | [Test/source/output] | [Decision] |

# Preflight and session information {-}

```{r preflight}
# Add every object required to reproduce the final table and figures.
required_objects <- c(
  "train_id", "test_id", "x_train", "x_test", "y_train", "y_test", "foldid"
)
missing_objects <- setdiff(required_objects, ls())
if (length(missing_objects)) {
  stop(
    "Missing required objects: ",
    paste(missing_objects, collapse = ", ")
  )
}
stopifnot(length(intersect(train_id, test_id)) == 0L)
```

```{r session-info}
sessionInfo()
```
