---
title: "Problem 2C"
output:
  pdf_document: default
---
# Setup helper functions
```{r setup}

source("utils.R")
library(car)
library(lars)
library(glmnet)

load("../data/tmpData.RData")
train_df <- analysis_data[train_id, ]
```

```{r lasso-fit}
lasso_cv <- fit_cv_glmnet(
    x = x_train,
    y = y_train,
    alpha = 1,
    foldid = foldid
)
plot(lasso_cv)

```
Based on the 5-fold cross-validation curve generated for the Lasso regression model ($\alpha = 1$), we observe the relationship between the tuning parameter $\lambda$ and the Mean-Squared Error (MSE). The top axis indicates the number of active (non-zero) predictors retained in the model at each level of penalization.

* The Cross-Validation Curve: The red dots represent the average CV-MSE across the 5 folds for each $\lambda$ value, with the gray error bars denoting one standard error above and below the mean. As $-\log(\lambda)$ increases (moving from left to right, meaning the penalty weakens), the MSE initially drops, reaches a minimum, and then slightly stabilizes as the model approaches the unpenalized OLS limit.

* Optimal $\lambda$ ($\lambda_{\text{min}}$): The right vertical dashed line marks `lambda.min`, the value that strictly minimizes the cross-validation MSE. At this optimal point for prediction accuracy ($-\log(\lambda) \approx 2.4$), the Lasso penalty retains 5 non-zero predictors*.

* Parsimonious $\lambda$ ($\lambda_{\text{1se}}$): The left vertical dashed line marks `lambda.1se`, selected by the one-standard-error rule. This value applies a stronger penalty ($-\log(\lambda) \approx 1.67$) to produce a sparser model. It retains only 3 non-zero predictors while keeping the error within one standard deviation of the absolute minimum MSE, offering a robust trade-off between predictive power and model simplicity.

```{r lasso-min-se}
lasso_cv$lambda.min

lasso_cv$lambda.1se
```
The exact penalty parameters determined by the cross-validation procedure are:

lambda.min: 0.0891. This is the penalty value that yields the lowest training cross-validation MSE (0.5790).

lambda.1se: 0.1875. Following the one-standard-error rule, this is the largest penalty parameter that keeps the cross-validation error within one standard deviation of the absolute minimum.


```{r lasso-nonzeros}
coef(lasso_cv, s="lambda.min")
count_nonzero(lasso_cv, s="lambda.min")
```

```{r lasso-nonzeros}
coef(lasso_cv, s="lambda.1se")
count_nonzero(lasso_cv, s="lambda.1se")

```

Comparing the active predictors under both tuning rules explicitly reveals Lasso's capability for exact variable selection:

Under lambda.min: The model retains 5 non-zero predictors (lcavol, lweight, lbph, svi, and pgg45). The remaining variables (age, lcp, and gleason) are shrunken completely to zero.

Under lambda.1se: The stronger penalty further simplifies the model by dropping lbph and pgg45. Only the 3 most statistically robust predictors are retained as non-zero: lcavol, lweight, svi.

```{r lasso-cv-error}
lasso_cv$cvm[
    lasso_cv$lambda==
    lasso_cv$lambda.min
]
```
The value 0.5790 represents the minimum cross-validation Mean-Squared Error (CV-MSE). 
It indicates the best predictive accuracy achieved by the Lasso model during the tuning phase across the 5 shared validation folds.

```{r lasso-coef-path}
lasso_fit <- glmnet(
    x_train,
    y_train,
    alpha=1,
    standardize=FALSE,
      
)
plot(lasso_fit)

```



