---
output:
  pdf_document: default
  html_document: default
---

--
title: "Problem 1C "
author: "Group 4"
date: "2026-07-14"
output: pdf_document
---
# Setup helper functions
```{r setup}

source("utils.R")
library(car)
# Use analysis_data[train_id, ] only. Add focused plots and summaries.
train_df <- analysis_data[train_id, ]
```

# 1C. Training-data exploration
## Pairwise correlation among predictors
```{r training-eda, eval=FALSE}
# Pairwise correlation among predictors
cor_matrix <- cor(
  train_df[, predictors],
  use = "complete.obs"
)

knitr::kable(
  round(cor_matrix, 2),
  caption = "Pairwise correlation matrix among predictors"
)

```

The pairwise correlation matrix was examined to identify potential relationships among predictors. Most predictors showed weak to moderate correlations, suggesting limited redundancy between features. However, some stronger positive correlations were observed, particularly between gleason and pgg45 (r = 0.78), lcavol and lcp (r = 0.68), and svi and lcp (r = 0.69). These relationships indicate that some predictors may share related clinical information. Therefore, Variance Inflation Factor (VIF) analysis was further performed to assess the overall multicollinearity effect in the regression model.


```{r training-eda, eval=FALSE}
# Response Distribution 
hist(y_train)
summary(y_train)
```
The distribution of the response variable (lpsa) was examined using a histogram and summary statistics. The response values range from -0.43 to 5.58, with a median of 2.57 and a mean of 2.48. The mean and median are relatively close, indicating that the distribution is approximately symmetric with no strong skewness. Although some observations appear at the lower and higher ends of the range, no severe outliers are evident from the overall distribution. Therefore, no transformation was applied to the response variable before model training.

```{r training-eda, eval=FALSE}
## Historgram
par(mfrow = c(2,4))

for (var in predictors) {
  hist(
    train_df[[var]],
    main = paste("Distribution of", var),
    xlab = var,
    breaks = 15
  )
}

par(mfrow = c(1,1))

## Boxplot
par(mfrow = c(2,4))

for (var in predictors) {
  boxplot(
    train_df[[var]],
    main = var
  )
}

par(mfrow = c(1,1))

```
The histograms show that most predictors have different distribution patterns. lcavol, lweight, and age are approximately symmetric, while lbph, lcp, and pgg45 exhibit right-skewed distributions with longer tails toward higher values. svi and gleason are discrete variables with highly concentrated values. These distribution characteristics are considered in later regression analysis and model diagnostics.

```{r training-eda, eval=FALSE}
# Correlation between predictors and response
cor_response <- cor(
  train_df[, predictors],
  train_df$lpsa
)

knitr::kable(
  round(cor_response, 2),
  caption = "Correlation between predictors and lpsa"
)
```

The correlation analysis indicates that all predictors have positive correlations with lpsa. lcavol shows the strongest relationship (r = 0.77), followed by lcp (r = 0.59) and svi (r = 0.57). Variables such as pgg45, gleason, and lweight have moderate correlations, while age and lbph show weak associations. Therefore, lcavol, lcp, and svi are likely to be important predictors in the multiple linear regression model.

```{r training-eda, eval=FALSE}
# Check multicollinearity using VIF
vif_model <- lm(
  lpsa ~ .,
  data = train_df[, c("lpsa", predictors)]
)

vif_values <- vif(vif_model)

knitr::kable(
  round(vif_values, 2),
  caption = "Variance Inflation Factor (VIF)"
)
```

There is no significant multicollinearity present in this model. Because all VIF values are safely below 5, the independent variables are not highly correlated with one another. The regression model's coefficient estimates are mathematically stable, and no variables need to be dropped or transformed for multicollinearity reasons.
```{r training-eda, eval=FALSE}
# Predictor-response plots
par(mfrow = c(2,4))

for (var in predictors) {
  plot(
    train_df[[var]],
    train_df$lpsa,
    xlab = var,
    ylab = "lpsa",
    main = paste(var, "vs lpsa"),
    pch = 19
  )

  abline(
    lm(train_df$lpsa ~ train_df[[var]]),
    col = "red"
  )
}

par(mfrow = c(1,1))
```
The provided scatter plots visualize the marginal relationships between the eight candidate predictors and the target variable, lpsa.

Strong Trends: lcavol demonstrates the clearest and most distinct positive linear correlation with lpsa, suggesting it will be a strong predictor in the model.

Categorical Patterns: Variables such as svi (binary) and gleason (ordinal) display expected clustered structures, with higher categories generally corresponding to higher lpsa values.

Weak Trends: age and lbph exhibit highly dispersed data clouds with very flat regression lines, indicating weak independent linear relationships with the target.

